Cross-relaxation in dense ensembles of NV centers and application to magnetometry \\ \medskip Relaxation croisées d'ensembles de centres NV et application à la magnétométrie

### 2022-10-22 Manuscrit Thèse/These.tex
FINITO LESPAGNOLO

While the NV physics can explain cross-relaxations between non-equivalent NV centers, it cannot explain cross-relaxation between equivalent NV centers. Nevertheless, as we will see in this section, CR between both equivalent and non-equivalent NV centers is observed for samples with high NV density. 
\subsection{NV-NV CR between non-equivalent NV centers}
\label{non_equi_valent_CR}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/NV-NV_non_equivalent}
\caption{a) Illustration of the four possible direction of the NV axis (four classes of NV centers) with a magnetic field along class 1 b) Representation of the polarization of two non equivalent spins, one aligned with the magnetic field (well polarized) and one misaligned with the magnetic field (unpolarized). c) Simulated frequencies for the $\ket{0} \to \ket{-1}$ and $\ket{0} \to \ket{+1}$ transitions when $\vec{B}\parallel [111]$ for the class aligned with B, and the 3 equivalent classes not aligned with B. d) Taken from \citep{armstrong2010nv}, PL change of an ensemble of NV centers as a magnetic field is scanned along the [111] axis.}
\label{non-equivalent NV-NV}
\end{figure}

In the absence of magnetic gradient, two NV centers can only be non-equivalent if the projection of the magnetic field upon their respective NV axis is different. Fig. \ref{non-equivalent NV-NV}-a) illustrates such a case: we label here the four possible NV axis orientation (or NV classes) 1 to 4 and we assume that the magnetic field is aligned with class 1. In this conditions, an NV center belong to class 1 will always be non-equivalent with NV centers from classes 2-4, as the projection of the magnetic field on their respective axis will never be the same. On the other hand, NV centers from the classes 2-4 are all equivalents, even though they do not necessary belong to the same class.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/NV-NV_equivalent}
\caption{CR between equivalent NV centers for sample ADM-15-1. (a) Representation of equivalent NV centers: either two NVs from the same class, or two classes with the same projected magnetic field. (b) Magnetic setup used for the experiment: a permanent magnet is used to apply a bias magnetic field, and an electromagnet is used to add a variable magnetic field. (c) Simulation showing the $\ket{0}\to\ket{-1}$ transition frequencies for the 4 classes of NV centers as a function of the scanned magnetic field. The transitions were computed based on ODMR spectra recorded for different magnetic field values. (d) Change in the PL of the NV centers ensemble as a function of the scanned magnetic field.}
\label{equivalent NV-NV}
\end{figure}

@@

\subsection{CR in an inhomogeneous NV bath}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/inhomogene_2_spins}
\caption{a) Schematics of the model: we consider a four level system consisting of 2 levels of 2 NV centers resonantly coupled through the flip-flop rate $\Gamma_{\rm dd}$. These two NV centers are each pumped in their respective $\ket{0}$ state via the green laser with a rate $\Gamma_{\rm las}$, and have their own relaxation rate $\Gamma_1$ and $\Gamma_1^*$ respectively. b) Average population of the two NV centers in the $\ket{0}$ state in the steady state as a function of $\Gamma_{\rm dd}$ and $\Gamma_1^*$. The steady state was computed by solving a classical rate equation where we fixed $\Gamma_{\rm las}=1000\ \rm Hz$ and $\Gamma_1=300\ \rm Hz$.}
\label{inhomogene}
\end{figure}
A likely explanation to the observed NV-NV CR is that the NV centers are not all equivalents.

Indeed, if we assume that, for instance, the relaxation rate $\Gamma_1=1/T_1$ is not strictly the same for each NV centers, but instead follows a certain distribution $\rho(\Gamma_1)$, then CR between NV centers with different $\Gamma_1$ can explain the observations in Fig. \ref{equivalent NV-NV}.

Fig. \ref{inhomogene}-a) and b) illustrates this case. Our model here consists of two (resonantly coupled) NV centers with two different relaxation rates $\Gamma_1$ and $\Gamma_1'$. To complete the model, we add the rate $\Gamma_{\rm las}$ which is the optical pumping rate from the $\ket{-1}$ to the $\ket{0}$ state of each spin, and $\Gamma_{\rm dd}$ the flip-flop rate between the two spins \footnote{We consider here a two level system instead of the full three levels of the spin 1. The choice of $\ket{-1}$ instead of $\ket{+1}$ is arbitrary.}. We only consider here the incoherent dynamics of the population, modeled with the rates reported on Fig. \ref{inhomogene}-a). Solving for the steady state of these rate equations, we can compute the final population in the $\ket{0}$ state for both spins.

@@

We can also notice that the influence of the flip-flop rate on $\rho_{00}$ increases when the difference in decay rates between NV$_1$ and NV$_2$ increases. As a result, we can suppose that a greater inhomogeneity between the NV centers would result in a stronger NV-NV cross-relaxation effect.

\subsection{Presentation of the fluctuator model}
\begin{figure}[h]
\centering
\includegraphics[width=.8\textwidth]{chapter 3/Figures/NV_vs_fluctuator}
\caption{Representation of a ``normal" NV center with spin depolarization rate $\Gamma_1 \ll \Gamma_{\rm las}$ and a fluctuator NV with spin depolarization rate $\gamma_f \gg \Gamma_{\rm las}$. Red circles represent the spin population in the steady state of each spin.}
\label{NV vs fluct}
\end{figure}
Choi et al. in \citep{choi2017depolarization} take this approach a step further by separating the NV centers into two groups: ``normal" NV centers with a phonon-limited lifetime outside of dipole-dipole interaction, and ``fluctuators" which are NV centers with an intrinsic, extremely fast relaxation mechanism leading to a new depolarization rate $\gamma_f$. Normal and fluctuator NV are represented in Fig. \ref{NV vs fluct}: in their model, Choi et al. assume that the fluctuator decay rate is fast enough compared to the laser polarization rate $\gamma_f \gg \Gamma_{\rm las}$ that the fluctuators are effectively always depolarized. Fluctuators are therefore similar to the dark spins we introduced in the last chapter.

@@

We will first decompose the dipole-dipole Hamiltonian as a product between radial and angular part:
\begin{equation}
\mathcal{H}_{\rm dd} \approx - \frac{J_0}{r^3} \left[(g+ih)(\ket{0,+1}\bra{+1,0}+\ket{0,-1}\bra{-1,0}+qS_z^1 S_z^2 \right] + h.c. ,
\end{equation}
where the expression of $J_0, g, h$ and $q$ is given in appendix \ref{Appendix eta}. $g, h$ and $q$ are dimensionless factors that are function of the relative radial positions and orientation of the two dipoles.

We then introduce the dimensionless number $\eta$ defined as:
\begin{equation}
\label{eq eta}
\eta^2=\frac{1}{3} (\abs{g}^2+\abs{h}^2) \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2},
\end{equation}
where $\gamma_f$ is the fluctuator decay rate for each possible channel between $\ket{0}, \ket{-1}$ and $\ket{+1}$. This number $\eta$ encapsulates both the angular dependency (through $g$ and $h$) and the resonance condition between the NV center and the fluctuator.

@@

We consider the situation pictured in Fig. \ref{NV + bain fluct}: the NV center is at the center of the frame and the position of each fluctuator is labeled $\vec{r}_i$. We consider that each fluctuator is an independent depolarization source, therefore the depolarization on a single NV center reads: 
\begin{equation}
\gamma=\sum_i \gamma_s(\vec{r}_i).
\end{equation}
We then want to compute the distribution $\rho(\gamma)$ defined as:
\begin{equation}
\rho(\gamma)=\int d\{r_i\} \rho(\{r_i\})\, \delta \left( \sum_i \gamma_s(\vec{r}_i) - \gamma \right).
\end{equation}
To do so, we need to determine the distribution of the fluctuators positions and orientations$\{r_i\}$. Assuming that they are homogeneously distributed in the bulk of the material, we find:
\begin{equation}
\rho(\gamma)=\frac{e^{-1/(4\gamma T)}}{\sqrt{4\pi \gamma^3 T}},
\end{equation}
where the time constant $T$ was introduced and is defined as:
\begin{equation}
\frac{1}{T}=\left(\frac{4\pi n_fJ_0\bar \eta}{3}\right)^2 \frac{\pi}{\gamma_f},
\label{eq 1/T}
\end{equation}
with $n$ the fluctuator density and $\bar \eta$ the averaged value of $\abs{\eta}$:

@@

We will start by separating the factor $\eta$ given in eq. (\ref{eq eta}) into a geometric and a spectral part:
\begin{align}
\eta^2&=G^2(\{\theta\}) R^2(\omega_f,\omega_{NV}), \\
G^2&=\frac{1}{3}\left(\abs{g}^2+\abs{h}^2 \right), \\ 
R^2(\omega_f,\omega_{NV}) &= \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2}.
\end{align}

@@

The computation of $\bar G$ will be further discussed in sec. \ref{sec quantitative T1} and in appendix \ref{delta gamma 1}. We will focus here on $\bar R$.

@@

In the case where $\omega^0_{NV}=\omega^0_{f}$, meaning that either the considered NVs and fluctuators are from the same class, or from two classes perfectly resonant, by substituting $\overline{R ^2}$ for $(\bar{R})^2$ (which again is only approximately valid for $2 \gamma_f \gg 1/T_2^*$) in eq. (\ref{eq 1/T}), we find that:
\begin{equation}
\label{eq 1/T avec T2*}
\frac{1}{T} \propto \frac{1}{2\gamma_f + \Gamma_f + \Gamma_{NV}}\cdot \frac{1}{1+\left(\frac{\omega^0_{NV}-\omega^0_{f}}{2 \gamma_f + \Gamma_f + \Gamma_{NV}}\right)^2}.
\end{equation}

@@

The second spin relaxation presented in Fig. \ref{stretch_or_not_stretch}-c) and d) was performed on sample ADM-15-2 which showed a great amount of dipole-dipole spin relaxation. To further enhance the cross-relaxation rate, the experiment was performed in zero external magnetic field so that every class of NV centers would be resonant. We can observe a relaxation profile that deviates significantly from tthe exponential profile, and which is nicely fitted by a stretched exponential. The $T_1$ value obtained from the fit is also an order of magnetite lower than the phonon-limited value, which is coherent with the fact that dipole-dipole interactions dominate the spin relaxation in this case.

@@

For many samples, the $T_1$ profile is neither fully exponential nor fully stretched exponential, because the relaxation time associated with CR, which we will call $T_1^{\rm dd}$, is of the same order of magnitude as the phonon limited lifetime $T_1^{\rm ph}$. Sometimes the same sample can go from mostly exponential to mostly stretched exponential depending on the number of NV-NV co-resonances, meaning that we have to include both aspects in the fitting formula.

@@

The fits used here are either purely exponential or purely stretched exponential, or a combination of both where the exponential lifetime $T_1^{\rm ph}$ was either left as a free parameter or fixed at a value $T_1^{\rm ph}=5\ \rm ms$. The values of the different fitting parameters used is reported in Table \ref{fitting table}.

We can see that neither the purely exponential nor stretched exponential fits can be satisfying for both measurements: the $B=0$ curve is poorly fitted by the exponential fit and the $B \neq 0$ curve is poorly fitted by the stretched exponential. The protocols that include both exponential and stretched lifetimes correctly fit both curves. In the case of Fig. \ref{various_fit_formulae}-d), we arbitrarily fixed $T_1^{\rm ph}=5\ \rm ms$ since this is the value we typically measure on samples with low NV density, and we expect that the phonon-limited exponential lifetime is not modified by the NV concentration.

@@

We should note however that the values of $T_1^{\rm dd}$ we obtain are not absolute. Indeed, in the case of Fig. \ref{various_fit_formulae}, the value of $T_1^{\rm ph}$ can be somewhat arbitrarily fixed between 3 and 6 ms with satisfying fits, and the resulting values $T_1^{\rm dd}$ can vary by almost an order of magnitude in the case of $B\neq0$. The values of $T_1^{\rm dd}$ obtained with this method have therefore to be compared to values from the same method and the same fixed value of $T_1^{\rm ph}$.

@@

Finally, another fitting procedure is shown in Fig. \ref{betas}. This time a single stretched exponential with an arbitrary stretch factor $\beta$ is used, and satisfyingly fits the data. Fig. \ref{betas}-b) shows the optimal $\beta$ parameter as a function of the external magnetic field. It confirms that the $T_1$ profile gets closer to a stretched exponential ($\beta=0.5$) when $B$ goes to 0. Even though this method also yields satisfying fits, it employs more free parameters ($\beta$ and $T_1$) than the method of \ref{various_fit_formulae}-d). We therefore decided to go with the later.

@@

\subsubsection{The fluctuators spectral response}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/fluctuator linewidth}
\caption{Measurement of the fluctuators linewidth on sample ADM-150-2 using the same setup as in Fig. \ref{equivalent NV-NV}. a) Simulation of the $\ket{0} \to \ket{-1}$ transition frequency for the four classes of NV centers and actual frequencies of the two central classes from ODMR measurements. b) Measurement of $T_1^{\rm dd}$ as a function of the splitting $\Delta \nu$ between the two central classes, fitted with a Lorentzian of half-width $\sigma=8.8 \pm 0.2\ \rm MHz$. The green dashed line correspond to the ODMR spectrum of a single class of NV centers, and the red dashed line correspond to the same spectrm scaled up by a factor of 2.}
\label{fluct linewidth}
\end{figure}

We now move to the second major prediction of the fluctuator model which is the spectral broadening of the fluctuators caused by the their relaxation rate $\gamma_f$. Indeed, following the arguments presented in \citep{choi2017depolarization}, the fluctuators homogeneous spectral response is a Lorentzian of half-width $2\gamma_f$. The inhomogeneous spectral response of the ensemble of fluctuators is then further broadened by the same electric, magnetic or strain noises than the NV centers, which is characterized by $1/T_2^*$.

@@

We can clearly see that the $\Gamma^{\rm dd}$ profile is much broader than the ODMR line of the NV centers, whereas in a standard CR process, we expect the decay rate to be proportional to the spectral overlap of the two classes \citep{hall2016detection}. If we assume that both NV classes spectral response are Lorentzian of same width $\sigma$, then the spectral overlap between the two classes is also a Lorentzian of width $2 \sigma$ \footnote{For other line profile, the overlap would be smaller than $2 \sigma$. For instances, the overlap of two Gaussians of width $\sigma$ is a Gaussian of width $\sqrt{2} \sigma$.}. Fig. \ref{fluct linewidth}-b) also shows the ODMR line scaled by a factor $2$, which is still significantly narrower than the $1/T_1^{\rm dd}$ line.

@@

\subsubsection{The fluctuators lifetime}
\label{sec fluctuators lifetime}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{chapter 3/Figures/deconvolution}
\caption{Blue curve $1/T_1^{\rm dd}$ data presented in Fig. \ref{fluct linewidth}. Red curve: deconvolution of the blue curve by the ODMR line shown on the same figure. Green curve: deconvolution of the red curve by the ODMR line. All three curves are fitted with Lorentzian of respective half widths $8.8 \pm 0.2$, $7.5 \pm 0.3$ and $6.4 \pm 0.4$ MHz (the precision is determined by the variance of the optimal fit parameters).}
\label{deconvolution}
\end{figure}

We now want to extract the fluctuators lifetime $T_1^f=1/\gamma_f$ to get further information on their potential nature.

The first way to do this, as was done for NV-P1 CR in \citep{hall2016detection}, is to deconvolve the $1/T_1^{\rm dd}$ data by the ODMR spectrum using a Wiener deconvolution algorithm. The result of the deconvolution is shown in Fig. \ref{deconvolution}: the data is deconvolved once to get the spectral response of the fluctuator (red curve), and a second time to get only the lifetime contribution of the fluctuator (green curve). Fitting with Lorentzians, we find a final value $\Gamma_2 =2\gamma_f= (2\pi) 6.4 \pm 0.4\ \rm MHz$ which corresponds to a fluctuator lifetime $T_1^f=\frac{1}{\gamma_f}=50 \pm 3\ \rm ns$\footnote{We define here the fluctuators lifetime $T_1^f$ as $\frac{1}{\gamma_f}$ for simplicity, but it should be noted that the actual lifetime of the fluctuators states is $\frac{1}{2\gamma_f}$ due to the competition between the two possible decay channels for each state (cf Fig. \ref{NV vs fluct}).}.

With $f=8.8 \pm 0.2\ \rm MHz$ and $\sqrt{2}f_G=4.3 \pm 0.2\ \rm MHz$, we find $f_L=6.5 \pm 0.3\ \rm MHz$ for a final lifetime value $T_1^f = 49 \pm 2\ \rm ns$, a value consistent with the one found with the deconvolution method.

Doing similar experiments and calculations, \citep{choi2017depolarization} found $\gamma_f=1/T_1^f=(2\pi) 3.3\ \rm MHz$ which gives them a fluctuator lifetime value $T_1^f = 48\ \rm ns$, a value very close to the ones we obtained above, despite using a completely different sample.

@@

The main tool to probe NV-NV CR is to use the co-resonances between the NV classes. While it is not possible to completely shut down the CR, because the NV centers from the same class are always resonant with each other, we can tune the density of resonant NV centers (and resonant fluctuators) by a factor of 4 by playing with the inter-class co-resonances. In this section, we will discuss the various scenarios where NV classes overlap, and quantify theoretically and experimentally the dipole-induced relaxation rate in each case.

@@

Fig. \ref{Carte} shows a map where we monitored the PL of sample CVD-pink while scanning the magnetic field angle around the [001] axis, thanks to a permanent magnet on a motorized goniometer. We have also reported the locus of $\vec{B}$ being in the planes [100], [010], [110] and [$1\bar 1 0$].

@@

Table \ref{T1 champ mag} shows these predicted values, along with experimental ones on sample ADM-15-3. The experimental values were obtained with the protocol described in Fig. \ref{various_fit_formulae}-d) with a double exponential fit where we fixed again $T_1^{\rm dd}=5\ \rm ms$.

@@

In conclusion, we have seen in this chapter that dense NV ensemble present ``anormal" CR between the NV centers, and we have shown how the fluctuator model developed by Choi et al. in \citep{choi2017depolarization} can explain this behavior. The fluctuator model not only explains the NV-NV CR, but also correctly predicts the stretched exponential lifetime profile. It also correctly predicts that the fluctuators have a broader spectral response than the other NV centers. Nevertheless, the model does not describe perfectly every aspects of NV-NV relaxation, and there still are many aspects of this problem, both theoretical and experimental, to explore.

@@

We should note that we observed large discrepancy in the zero field depolarization among the samples that we studied, even from samples coming from the same batch.

@@

>>>>

While the single NV physics can explain cross-relaxations between non-equivalent NV centers, it cannot explain cross-relaxation between equivalent NV centers. Nevertheless, as we will see in this section, CR between both equivalent and non-equivalent NV centers is observed for samples with high NV density. 
\subsection{NV-NV CR between non-equivalent NV centers}
\label{non_equi_valent_CR}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/NV-NV_non_equivalent}
\caption{a) Illustration of the four possible direction of the NV axis (four classes of NV centers) with a magnetic field along class 1 b) Representation of the polarization of two non equivalent spins, one aligned with the magnetic field (well polarized) and one misaligned with the magnetic field (unpolarized). c) Simulated frequencies for the $\ket{0} \to \ket{-1}$ and $\ket{0} \to \ket{+1}$ transitions when $\vec{B}\parallel [111]$ for the class aligned with B, and the 3 equivalent classes not aligned with B. d) Taken from \citep{armstrong2010nv}, PL change of an ensemble of NV centers as a magnetic field is scanned along the [111] axis.}
\label{non-equivalent NV-NV}
\end{figure}

In the absence of magnetic gradient, two NV centers can only be non-equivalent if the projection of the magnetic field upon their respective NV axis is different. Fig. \ref{non-equivalent NV-NV}-a) illustrates such a case: we label here the four possible NV axis orientation (or NV classes) 1 to 4 and we assume that the magnetic field is aligned with class 1. In this conditions, an NV center belonging to class 1 will always be non-equivalent with NV centers from classes 2-4, as the projection of the magnetic field on their respective axis will never be the same. On the other hand, NV centers from the classes 2-4 are all equivalents, even though they do not necessary belong to the same class.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/NV-NV_equivalent}
\caption{CR between equivalent NV centers for sample ADM-15-1 (see sample details in appendix \ref{Appendix samples}). (a) Representation of equivalent NV centers: either two NVs from the same class, or two classes with the same projected magnetic field. (b) Magnetic setup used for the experiment: a permanent magnet is used to apply a bias magnetic field, and an electromagnet is used to add a variable magnetic field. (c) Simulation showing the $\ket{0}\to\ket{-1}$ transition frequencies for the 4 classes of NV centers as a function of the scanned magnetic field. The transitions were computed based on ODMR spectra recorded for different magnetic field values. (d) Change in the PL of the NV centers ensemble as a function of the scanned magnetic field.}
\label{equivalent NV-NV}
\end{figure}

@@

\subsection{CR in an inhomogeneous NV bath}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/inhomogene_2_spins}
\caption{a) Schematics of the model: we consider a four level system consisting of 2 levels of 2 NV centers resonantly coupled through the flip-flop rate $\Gamma_{\rm dd}$. These two NV centers are each pumped in their respective $\ket{0}$ state via the green laser with a rate $\Gamma_{\rm las}$, and have their own relaxation rate $\Gamma_1$ and $\Gamma_1^*$ respectively. b) Average population of the two NV centers in the $\ket{0}$ state in the steady state as a function of $\Gamma_{\rm dd}$ and $\Gamma_1^*$. The steady state was computed by solving for a classical rate equation where we fixed $\Gamma_{\rm las}=1000\ \rm Hz$ and $\Gamma_1=300\ \rm Hz$.}
\label{inhomogene}
\end{figure}
A likely explanation to the observed NV-NV CR is that the NV centers are not all equivalents.

Indeed, if we assume that, for instance, the relaxation rate $\Gamma_1=1/T_1$ is not strictly the same for each NV centers, but instead follows a certain distribution $\rho(\Gamma_1)$, then CR between NV centers with different $\Gamma_1$ can explain the observations in Fig. \ref{equivalent NV-NV}-d).

Fig. \ref{inhomogene}-a) and b) illustrates this case. Our model here consists of two (resonantly coupled) NV centers with two different relaxation rates $\Gamma_1$ and $\Gamma_1'$. To complete the model, we add the rate $\Gamma_{\rm las}$ which is the optical pumping rate from the $\ket{-1}$ to the $\ket{0}$ state of each spin, and $\Gamma_{\rm dd}$ the flip-flop rate between the two spins \footnote{We only consider two levels for each spin instead of the full three levels of the spin 1. The choice of $\ket{-1}$ instead of $\ket{+1}$ is arbitrary.}. We only consider here the incoherent dynamics of the population, modeled with the rates reported on Fig. \ref{inhomogene}-a). Solving for the steady state of these rate equations, we can compute the final population in the $\ket{0}$ state for both spins.

@@

We can also notice that the influence of the flip-flop rate on $\rho_{00}$ increases when the difference in decay rates between NV$_1$ and NV$_2$ increases. As a result, we conclude that a greater inhomogeneity between the NV centers results in a stronger NV-NV cross-relaxation effect.

\subsection{Presentation of the fluctuator model}
\begin{figure}[h]
\centering
\includegraphics[width=.8\textwidth]{chapter 3/Figures/NV_vs_fluctuator}
\caption{Representation of a ``normal" NV center with spin depolarization rate $\Gamma_1 \ll \Gamma_{\rm las}$ and a fluctuator NV with spin depolarization rate $\gamma_f \gg \Gamma_{\rm las}$. Red circles represent the spin population in the steady state of each spin.}
\label{NV vs fluct}
\end{figure}
Choi et al. in \citep{choi2017depolarization} take this approach a step further by separating the NV centers into two groups: ``normal" NV centers with a phonon-limited lifetime (not taking into account the dipole-dipole interaction), and ``fluctuators" which are NV centers with an intrinsic, extremely fast relaxation mechanism leading to a new depolarization rate $\gamma_f$. Normal and fluctuator NV are represented in Fig. \ref{NV vs fluct}: in their model, Choi et al. assume that the fluctuator decay rate is fast enough compared to the laser polarization rate $\gamma_f \gg \Gamma_{\rm las}$ that the fluctuators are effectively always depolarized. Fluctuators are therefore similar to the dark spins we introduced in the last chapter.

@@

We will first decompose the dipole-dipole Hamiltonian as a product between radial and angular part:
\begin{equation}
\mathcal{H}_{\rm dd} \approx - \frac{J_0}{r^3} \left[(g+ih)(\ket{0,+1}\bra{+1,0}+\ket{0,-1}\bra{-1,0}+qS_z^1 S_z^2 \right] + h.c. ,
\end{equation}
where the expression of $J_0, g, h$ and $q$ are given in appendix \ref{Appendix eta}. $g, h$ and $q$ are dimensionless factors that are function of the relative radial positions and orientation of the two dipoles.

We then introduce the dimensionless number $\eta$ defined as:
\begin{equation}
\label{eq eta}
\eta^2=\frac{1}{3} (\abs{g}^2+\abs{h}^2) \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2},
\end{equation}
where $\gamma_f$ is the fluctuator decay rate for each possible channel between $\ket{0}, \ket{-1}$ and $\ket{+1}$ (cf Fig. \ref{NV vs fluct}). This number $\eta$ encapsulates both the angular dependency (through $g$ and $h$) and the resonance condition between the NV center and the fluctuator.

@@

We consider the situation pictured in Fig. \ref{NV + bain fluct}: the NV center is at the center of the frame of reference and the position of each fluctuator is labeled $\vec{r}_i$. We consider that each fluctuator is an independent depolarization source, therefore the depolarization on a single NV center reads: 
\begin{equation}
\gamma=\sum_i \gamma_s(\vec{r}_i).
\end{equation}
We then want to compute the distribution $\rho(\gamma)$ defined as:
\begin{equation}
\rho(\gamma)=\int d\{r_i\} \rho(\{r_i\})\, \delta \left( \sum_i \gamma_s(\vec{r}_i) - \gamma \right).
\end{equation}
To do so, we need to determine the distribution of the fluctuators positions and orientations$\{r_i\}$. Assuming that they are homogeneously distributed in the bulk of the material, we find:
\begin{equation}
\rho(\gamma)=\frac{e^{-1/(4\gamma T)}}{\sqrt{4\pi \gamma^3 T}},
\end{equation}
where the time constant $T$ was introduced and is defined as:
\begin{equation}
\frac{1}{T}=\left(\frac{4\pi n_fJ_0\bar \eta}{3}\right)^2 \frac{\pi}{\gamma_f},
\label{eq 1/T}
\end{equation}
with $n$ the fluctuator density and $\bar \eta$ the averaged value of $\abs{\eta}$:

@@

We will start by rewriting the factor $\eta$ given in eq. (\ref{eq eta}) as the product of a geometric and spectral part:
\begin{align}
\eta^2&=G^2(\{\theta\}) R^2(\omega_f,\omega_{NV}), \\
G^2&=\frac{1}{3}\left(\abs{g}^2+\abs{h}^2 \right), \\ 
R^2(\omega_f,\omega_{NV}) &= \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2}.
\end{align}

@@

The computation of $\bar G$ will be further discussed in sec. \ref{sec quantitative T1} and in appendix \ref{Appendix eta}. We will focus here on $\bar R$.

@@

By substituting $\overline{R ^2}$ for $(\bar{R})^2$ (which again is only approximately valid for $2 \gamma_f \gg 1/T_2^*$) in eq. (\ref{eq 1/T}), we find that:
\begin{equation}
\label{eq 1/T avec T2*}
\frac{1}{T} \propto \frac{1}{2\gamma_f + \Gamma_f + \Gamma_{NV}}\cdot \frac{1}{1+\left(\frac{\omega^0_{NV}-\omega^0_{f}}{2 \gamma_f + \Gamma_f + \Gamma_{NV}}\right)^2}.
\end{equation}

@@

The second spin relaxation presented in Fig. \ref{stretch_or_not_stretch}-c) and d) was performed on sample ADM-15-2 which showed a great amount of dipole-dipole spin relaxation. To further enhance the cross-relaxation rate, the experiment was performed in zero external magnetic field so that every class of NV centers would be resonant. We can observe a relaxation profile that deviates significantly from the exponential profile, and which is nicely fitted by a stretched exponential. The $T_1$ value obtained from the fit is also an order of magnetite lower than the phonon-limited value, which is coherent with the fact that dipole-dipole interactions dominate the spin relaxation in this case.

@@

For many samples, the $T_1$ profile is neither fully exponential nor fully stretched exponential, because the relaxation time associated with CR, which we will call $T_1^{\rm dd}$, is of the same order of magnitude as the phonon limited lifetime $T_1^{\rm ph}$. Some samples have both predominantly exponential and predominantly stretched exponential profile depending on the number of NV-NV co-resonances. We therefore have to include both aspects in the fitting formula.

@@

The fit functions investigated here are: pure exponential, stretched exponential, or a combination of both where the exponential lifetime $T_1^{\rm ph}$ was either left as a free parameter or fixed at a value $T_1^{\rm ph}=5\ \rm ms$. The values of the different fitting parameters used is reported in Table \ref{fitting table}.

We can see that neither the pure exponential nor stretched exponential can satisfyingly fit both measurements: the $B=0$ curve is poorly fitted by the exponential fit and the $B \neq 0$ curve is poorly fitted by the stretched exponential. The protocols that include both exponential and stretched lifetimes correctly fit both curves. In the case of Fig. \ref{various_fit_formulae}-d), we arbitrarily fixed $T_1^{\rm ph}=5\ \rm ms$ since this is the value we typically measure on samples with low NV density, and we expect that the phonon-limited exponential lifetime is not modified by the NV concentration.

@@

We should note however that the values of $T_1^{\rm dd}$ we obtain are not absolute. Indeed, in the case of Fig. \ref{various_fit_formulae}, the value of $T_1^{\rm ph}$ can be somewhat arbitrarily fixed between 3 and 6 ms, and the resulting values $T_1^{\rm dd}$ can vary by almost an order of magnitude in the case of $B\neq0$. The values of $T_1^{\rm dd}$ obtained with this method have therefore to be compared to values from the same method and the same fixed value of $T_1^{\rm ph}$.

@@

Finally, another fitting procedure is shown in Fig. \ref{betas}. This time a single stretched exponential with an arbitrary stretch factor $\beta$ is used, and satisfyingly fits the data. Fig. \ref{betas}-b) shows the optimal $\beta$ parameter as a function of the external magnetic field. It confirms that the $T_1$ profile gets closer to a stretched exponential ($\beta=0.5$) when $B$ goes to 0. Even though this method also yields satisfying fits, it employs more free parameters ($\beta$ and $T_1$) than the method of \ref{various_fit_formulae}-d). We therefore decided to work with the later.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{chapter 3/Figures/fluctuator linewidth}
\caption{Measurement of the fluctuators linewidth on sample ADM-150-2 using the same setup as in Fig. \ref{equivalent NV-NV}. a) Simulation of the $\ket{0} \to \ket{-1}$ transition frequency for the four classes of NV centers and actual frequencies of the two central classes from ODMR measurements. b) Measurement of $T_1^{\rm dd}$ as a function of the splitting $\Delta \nu$ between the two central classes, fitted with a Lorentzian of half-width $\sigma=8.8 \pm 0.2\ \rm MHz$. The green dashed line correspond to the ODMR spectrum of a single class of NV centers, and the red dashed line correspond to the same spectrum scaled up by a factor of 2.}
\label{fluct linewidth}
\end{figure}

We now move to the second major prediction of the fluctuator model which is the spectral broadening of the fluctuators caused by the their relaxation rate $\gamma_f$. Indeed, following the arguments presented in \citep{choi2017depolarization}, the inhomogeneous spectral response of the fluctuators is broadened by a Lorentzian of half-width $2\gamma_f$ on top of the ``normal" $T_2^*$ broadening resulting from magnetic, electric or strain noise.

@@

We can clearly see that the $\Gamma_1^{\rm dd}$ profile is much broader than the ODMR line of the NV centers, whereas in a standard CR process, we expect the decay rate to be proportional to the spectral overlap of the two classes \citep{hall2016detection}. If we assume that both NV classes spectral response are Lorentzian of same width $\sigma$, then the spectral overlap between the two classes is also a Lorentzian of width $2 \sigma$ \footnote{For other line profile, the overlap would be smaller than $2 \sigma$. For instances, the overlap of two Gaussians of width $\sigma$ is a Gaussian of width $\sqrt{2} \sigma$.}. Fig. \ref{fluct linewidth}-b) also shows the ODMR line scaled by a factor $2$, which is still significantly narrower than the $1/T_1^{\rm dd}$ line.

@@

We proceed to extract the value of $\gamma_f$ from the data presented in Fig. \ref{fluct linewidth} by two methods: one involves the deconvolution of the $\Gamma_1^{\rm dd}$ profile by the ODMR profile, and the other involves approximations with pseudo-Voigt profiles. Both methods are detailed in appendix \ref{Appendix fluct T1} and result in $\gamma_f$ values of $(2\pi) 3.3 \pm 0.2\ \rm MHz$ and $(2\pi) 3.2 \pm 0.2\ \rm MHz$ respectively, which corresponds to a fluctuator spin lifetime $T_1^f=\frac{1}{3 \gamma_f}\approx 17\ \rm ns$\footnote{The expression of $T_1^f$ as a function of $\gamma_f$ is found by solving for the time dependent solution of the rate equations associated with the fluctuator model shown in Fig. \ref{NV vs fluct}.}.

It is worth noting that that \citep{choi2017depolarization} found on their sample $\gamma_f=(2\pi) 3.3\ \rm MHz$ by performing similar experiments, a value very close to the ones we obtained above.

@@

The main tool to probe NV-NV CR is to use the co-resonances between the NV classes. While it is not possible to completely shut down the CR, because NV centers and fluctuators from the same class are always resonant with each other, we can tune the density of resonant NV centers and fluctuators by a factor of 4 by playing with the inter-class co-resonances. In this section, we will discuss the various scenarios where NV classes overlap, and quantify theoretically and experimentally the dipole-induced relaxation rate in each case.

@@

Fig. \ref{Carte} shows a map where we monitored the PL of sample CVD-pink while scanning the magnetic field angle around the [001] axis, thanks to a permanent magnet mounted on a motorized goniometer. We have also reported the locus of $\vec{B}$ being in the planes [100], [010], [110] and [$1\bar 1 0$].

@@

Table \ref{T1 champ mag} shows these predicted values, along with experimental ones measured on sample ADM-15-3 (see sample details in appendix \ref{Appendix samples}). The experimental values were obtained with the protocol described in Fig. \ref{various_fit_formulae}-d) with a double exponential fit where we fixed again $T_1^{\rm dd}=5\ \rm ms$.

@@

In conclusion, we have seen in this chapter that dense NV ensemble present ``anormal" CR between the NV centers, and we have shown how the fluctuator model developed by Choi et al. in \citep{choi2017depolarization} can explain this behavior. The fluctuator model not only explains the NV-NV CR, but also correctly predicts the stretched exponential lifetime profile. It also correctly predicts that the fluctuators have a broader spectral response than the other NV centers. Nevertheless, the model does not describe perfectly every aspects of NV-NV relaxation.

@@

We should note that we observed large discrepancies in the zero field depolarization among the samples that we studied, even from samples coming from the same batch.

@@

\chapter{Extraction of the fluctuator depolarization rate $\gamma_f$}
